Estimating initial state and quark-gluon plasma medium properties \\ using a hybrid model with nucleon substructure \\ calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV

### 2018-07-21 paper/substructure.tex
Fix some latex formatting issues

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{multirow}

\usepackage[inline]{enumitem}
\usepackage{lipsum}

@@

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\sigmaf}{\sigma_\mathrm{fluct}}
\newcommand{\X}{\chi_\mathrm{struct}}
\newcommand{\taufs}{\tau_\mathrm{fs}}
\newcommand{\dmin}{d_\mathrm{min}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

\renewcommand\labelitemi{$\vcenter{\hbox{\scriptsize$\bullet$}}$}

@@

% Convenient figure macro.  Usage:
%
%   \fig[placement specifier = t]{filename}{caption}
%
% This creates a figure environment, includes the given filename as graphics,
% puts the given caption below the graphics, and labels it 'fig:filename'.
% The optional placement specifier defaults to 't' and is passed directly to
% the figure environment.
% Use \fig* to make a figure* environment, i.e. a wide figure.
\usepackage{xparse}
\NewDocumentCommand\fig{sO{t}mm}{
  \begin{figure\IfBooleanT{#1}{*}}[#2]
    \includegraphics{#3}
    \caption{\label{fig:#3}#4}
  \end{figure\IfBooleanT{#1}{*}}
}

@@

\title{
  Estimating initial state and quark-gluon plasma medium properties\\
  using a hybrid model with nucleon substructure\\
  calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV
}

@@

Most of the calibration observables listed in Tbl.~\ref{tab:observables} are calculated as a function of collision centrality, where centrality is defined using some measure of the underlying event activity, e.g.\ the charged particle yield in a given rapidity window.
When calculating these observables, we generate $\mathcal{O}(10^4)$ minimum-bias events at each design point and divide the events into centrality bins using the charged particle yield at midrapidity, similar to the procedure used by experiment.

@@

For the correlation structure between different observable bins, we assert a simple Gaussian form
\begin{equation}
  \label{eq:corr}
  \rho_{ij}^\text{sys} = \exp \left[ -\frac{1}{2} \left(\frac{b_i - b_j}{l} \right)^2 \right],
\end{equation}

where $b_i$ and $b_j$ are the midpoints of two observable bins of a single type (centrality or relative multiplicity) and $l$ is a correlation length which describes how quickly the observable bins decorrelate as the distance between the bins increases.
We use centrality correlation lengths ${l=100\%}$ for all Pb-Pb observables and ${l=30\%}$ for the $p$-Pb charged particle yield $d\nch/d\eta$.
The $p$-Pb mean $p_T$ and flow observables, meanwhile, use relative multiplicity bins $\nch / \langle \nch \rangle$ and $\ntrk / \langle \ntrk \rangle$ so we use a much smaller correlation length $l=5$ to reflect the smaller range of bin values.
We show, as an example, the resulting experimental correlation matrix $\text{corr}(\Sigma_e)$ for the Pb-Pb system on the right side of Fig.~\ref{fig:correlation}.
The correlation matrix is block diagonal, with each block representing the correlations within a single class of observable; in this case, the experimental yield $d\nch/d\eta$ and flows $\vnk{n}{2}$.

@@

The principal components are reported in order of explained variance, with the first principal component vector $\mathbf{v}_1$ accounting for the most variance in $\tilde{Y}$, and the last principal component vector $\mathbf{v}_m$ accounting for the least.
We then train a set of independent GPE's,
\begin{align}
  \label{eq:gpe}
  z_1 &= \text{gp}_1(\x), \nonumber \\
  z_2 &= \text{gp}_2(\x), \nonumber \\
  \vdots & \nonumber\\
  z_k &= \text{gp}_k(\x),
\end{align}
to predict the principal components (coefficients) $(z_1, \dots, z_k)$ of the first $k$ principal component vectors as a function of the model parameters $\x$ which vary across the design $X$.
We use $k=7$ principal components when emulating the $p$-Pb system and $k=8$ principal components when emulating Pb-Pb, chosen to describe 99.5\% of the total observed variance of each system.

@@

\onecolumngrid
{\centering\includegraphics{observables_pbpb}\par}
\captionof{figure}{Simulated observables compared to experimental data for Pb-Pb collisions at $\sqrts=5.02$~TeV. Top row: explicit model calculations (no emulator) for each of the 500 design points; bottom row: emulator predictions for $n=100$ random samples drawn from the posterior. Points with error bars are experimental data from ALICE with statistical and systematic errors added in quadrature \note.}
\vspace{.5cm}
{\centering\includegraphics{observables_ppb}\par}
\captionof{figure}{Same as Fig.~\ref{fig:obs_pbpb} but for $p$-Pb collisions at $\sqrts=5.02$~TeV. Note that multiplicity bins are used for mean $p_T$ and flow cumulant observables to match the bins used by experiment. Experimental data are from ALICE and CMS \note.}
\twocolumngrid

@@

We then perform Markov-chain Monte Carlo (MCMC) importance sampling on the posterior distribution in Eqn.~\eqref{eq:bayes} to draw random samples for $P(H_\x | E)$, the estimate of the true model parameters given the model and the experimental data.
We use an affine-invariant ensemble sampler which uses a large ensemble of interdependent walkers \note\ and allow the MCMC chain to ``burn-in'' before generating $\mathcal{O}(10^7)$ posterior samples.

@@

{
\onecolumngrid
\begin{figure}
  \makebox[\textwidth]{
      \includegraphics[width=.95\paperwidth]{posterior}
  }
\end{figure}
\captionof{figure}{Bayesian posterior distribution of the model input parameters. The diagonal panels show the marginalized distributions of individual model parameters, while off-diagonal panels show the joint distributions for pairs of model parameters, visualizing their correlations. The marginalized distribution medians and 90\% credible intervals are annotated along the diagonal.}
\clearpage
\twocolumngrid
}

@@

\onecolumngrid
{\centering\includegraphics{observables_map}\par}
\captionof{figure}{Model calculations using the \emph{maximum a posterior} (MAP) parameters compared to experiment. Colored lines are model calculations for $p$-Pb collisions (left) and Pb-Pb collisions (right) at $\sqrts=5.02$~TeV. Points with error bars are the experimental data with statistical uncertainties, and gray bands their corresponding systematic uncertainties, from ALICE and CMS \note. The ratio axes show the ratio of model over data where where available with gray bands indicating $\pm 10\%$.}
\twocolumngrid

@@

\section{Emulator validation}
\label{app:validation}
\fig*[t]{validation_example}{
  Placeholder
}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}

\usepackage[inline]{enumitem}

@@

\newcommand{\trento}{T\raisebox{-0.5ex}{R}ENTo}
\newcommand{\sqrts}{\sqrt{s_{NN}}}
\newcommand{\nch}{N_\text{ch}}
\newcommand{\ntrk}{N_\text{trk}^\text{offline}}
\newcommand{\sigmaf}{\sigma_\mathrm{fluct}}
\newcommand{\X}{\chi_\mathrm{struct}}
\newcommand{\taufs}{\tau_\mathrm{fs}}
\newcommand{\dmin}{d_\mathrm{min}}
\newcommand{\T}{\tilde{T}}
\newcommand{\x}{\mathbf{x}}
\newcommand{\y}{\mathbf{y}}
\newcommand{\z}{\mathbf{z}}
\newcommand{\vnk}[2]{v_#1\{#2\}}
\newcommand{\paddedhline}{\noalign{\smallskip}\hline\noalign{\smallskip}}
\newcommand{\order}[1]{$\mathcal O(10^{#1})$}
\newcommand{\note}{\textcolor{theblue}{[?]}}

% Reduce the size of the itemize bullets
\renewcommand\labelitemi{$\vcenter{\hbox{\scriptsize$\bullet$}}$}

% Create a temporary full-page single column environment
\newenvironment{fullpage}{\onecolumngrid}{\clearpage\twocolumngrid}

@@

@@

\title{
  Estimating initial state and quark-gluon plasma medium properties \\
  using a hybrid model with nucleon substructure \\
  calibrated to p-Pb and Pb-Pb collisions at \texorpdfstring{$\mathbf{\sqrts=5.02}$}{}~TeV
}

@@

Most of the calibration observables listed in Tbl.~\ref{tab:observables} are calculated as a function of collision centrality, where centrality is defined using some measure of the underlying event activity, e.g.\ the charged particle yield in a given rapidity window.
When calculating these observables, we generate \order{10^4} minimum-bias events at each design point and divide the events into centrality bins using the charged particle yield at midrapidity, similar to the procedure used by experiment.

@@

For the correlation structure between different observable bins, we assert a simple Gaussian form
\begin{equation}
  \label{eq:corr}
  \rho_{ij}^\text{sys} = \exp \left[ -\frac{1}{2} \left(\frac{b_i - b_j}{l} \right)^2 \right],
\end{equation}
where $b_i$ and $b_j$ are the midpoints of two observable bins of a single type (centrality or relative multiplicity) and $l$ is a correlation length which describes how quickly the observable bins decorrelate as the distance between the bins increases.
We use centrality correlation lengths ${l=100\%}$ for all Pb-Pb observables and ${l=30\%}$ for the $p$-Pb charged particle yield $d\nch/d\eta$.
The $p$-Pb mean $p_T$ and flow observables, meanwhile, use relative multiplicity bins $\nch / \langle \nch \rangle$ and $\ntrk / \langle \ntrk \rangle$ so we use a much smaller correlation length $l=5$ to reflect the smaller range of bin values.
We show, as an example, the resulting experimental correlation matrix $\text{corr}(\Sigma_e)$ for the Pb-Pb system on the right side of Fig.~\ref{fig:correlation}.
The correlation matrix is block diagonal, with each block representing the correlations within a single class of observable; in this case, the experimental yield $d\nch/d\eta$ and flows $\vnk{n}{2}$.

@@

The principal components are reported in order of explained variance, with the first principal component vector $\mathbf{v}_1$ accounting for the most variance in $\tilde{Y}$, and the last principal component vector $\mathbf{v}_m$ accounting for the least.
We then train a set of independent GPE's,
\begin{align}
  \label{eq:gpe}
  z_1 &= \text{gp}_1(\x), \nonumber \\
  z_2 &= \text{gp}_2(\x), \nonumber \\
  \vdots & \nonumber \\
  z_k &= \text{gp}_k(\x),
\end{align}
to predict the principal components (coefficients) $(z_1, \dots, z_k)$ of the first $k$ principal component vectors as a function of the model parameters $\x$ which vary across the design $X$.
We use $k=7$ principal components when emulating the $p$-Pb system and $k=8$ principal components when emulating Pb-Pb, chosen to describe 99.5\% of the total observed variance of each system.

@@

\begin{fullpage}
{\centering\includegraphics{observables_pbpb}\par}
\captionof{figure}{Simulated observables compared to experimental data for Pb-Pb collisions at $\sqrts=5.02$~TeV. Top row: explicit model calculations (no emulator) for each of the 500 design points; bottom row: emulator predictions for $n=100$ random samples drawn from the posterior. Points with error bars are experimental data from ALICE with statistical and systematic errors added in quadrature \note.}
\vspace{.5cm}
{\centering\includegraphics{observables_ppb}\par}
\captionof{figure}{Same as Fig.~\ref{fig:obs_pbpb} but for $p$-Pb collisions at $\sqrts=5.02$~TeV. Note that multiplicity bins are used for mean $p_T$ and flow cumulant observables to match the bins used by experiment. Experimental data are from ALICE and CMS \note.}
\end{fullpage}

@@

We then perform Markov-chain Monte Carlo (MCMC) importance sampling on the posterior distribution in Eqn.~\eqref{eq:bayes} to draw random samples for $P(H_\x | E)$, the estimate of the true model parameters given the model and the experimental data.
We use an affine-invariant ensemble sampler which uses a large ensemble of interdependent walkers \note\ and allow the MCMC chain to ``burn-in'' before generating \order{10^7} posterior samples.

@@

\begin{fullpage}
\begin{figure}
  \makebox[\textwidth]{
      \includegraphics[width=.95\paperwidth]{posterior}
  }
\end{figure}
\captionof{figure}{Bayesian posterior distribution of the model input parameters. The diagonal panels show the marginalized distributions of individual model parameters, while off-diagonal panels show the joint distributions for pairs of model parameters, visualizing their correlations. The marginalized distribution medians and 90\% credible intervals are annotated along the diagonal.}
\end{fullpage}

@@

\begin{fullpage}
{\centering\includegraphics{observables_map}\par}
\captionof{figure}{Model calculations using the \emph{maximum a posterior} (MAP) parameters compared to experiment. Colored lines are model calculations for $p$-Pb collisions (left) and Pb-Pb collisions (right) at $\sqrts=5.02$~TeV. Points with error bars are the experimental data with statistical uncertainties, and gray bands their corresponding systematic uncertainties, from ALICE and CMS \note. The ratio axes show the ratio of model over data where where available with gray bands indicating $\pm 10\%$.}
\end{fullpage}

@@

\section{Emulator validation}
\label{app:validation}
\begin{figure*}
  \includegraphics{validation_example}
\end{figure*}
